---
title: "Parallel"
author: "Kondwani Kajera Mughogho"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 90, tidy = TRUE, warning = FALSE, message = FALSE)
```

```{r, error=TRUE}
ptm <- proc.time() # time the analysis

library(lsasim)
library(mvtnorm)
library(mirt)
```

This is blogpost shows how to use the R package parallel. I will also show how much you save up on time by parallelizing your code. 

First of all, I will specify my test characteristics. In this case, I will simulate a 45 item test which will be administered to 3000 examinees. The test simulations will be conducted over 3 replications. 
```{r cars}

n_examinees    <- 2000 # number of examinees
total_items    <- 30 # total number of items
no_reps        <- 12 # number of replications

```

I will then proceed on to generating my item parameters using R package lsasim. These item parameters will be from a 2 parameter-logistic model. In this case
````{r lsasim, error = TRUE}
items <- list() # specify an empty list for item parameters
set.seed( round(5653) ) # seed for generated item parameters
items <- lsasim::item_gen(n_2pl = total_items , 
                          b_bounds = c(-2, 2), # boundeies for the item difficulty parameters
                          a_bounds = c(0.75, 1.25) # bounderies for the item discrimination parameters
                          ) 

#--- combined sub-test items into a single data frame
test_items <- items # save item parameters 

````
In this case, it is after generating the item parameters, I may then add everything that I intend on manipulate into a function. It is this function that will be distributed to parallel nodes. Note that in the function statement, I include all components of the simulation that are to fixed. These include: items, test_items, and total_items. One other key thing to note is that, the packages that will be used in your simulation should be loaded in the function as well. 

```{r}
#--- determine replications
replications <- vector("list", no_reps) 
for (rep in 1:no_reps) {
  replications[[rep]] <- rep
}

myFunction <- function(X, items, test_items, 
                       n_examinees, total_items, verbose = TRUE) {
  # install packages
  library(mirt)
  library(mvtnorm)
  library(lsasim)

  # specify the number of replications
  kk <- X
  
  #------------------------------------------------------------------------------#
  ## Generate thetas 
  #------------------------------------------------------------------------------#
  theta <- rnorm(n_examinees, mean = 0, sd = 1)

  #------------------------------------------------------------------------------#
  # Generate item responses 
  #------------------------------------------------------------------------------#
  resp <- data.frame(matrix(NA, nrow= n_examinees, ncol = total_items))
  colnames(resp) <- paste0("i.", 1:total_items)
  
  # assign items to block
  block_bk1 <- lsasim::block_design(n_blocks = 1, 
                                    item_parameters = items)
  
  #assign block to booklet
  book_bk1 <- lsasim::booklet_design(item_block_assignment = 
                                     block_bk1$block_assignment,
                                     book_design = matrix(1))
  #assign booklet to subjects
  book_samp <- lsasim::booklet_sample(n_subj = n_examinees, 
                                      book_item_design = book_bk1, 
                                      book_prob = NULL)
  
  # generate item responses 
  cog <- lsasim::response_gen(subject = book_samp$subject, 
                              item = book_samp$item, 
                              theta = theta, 
                              b_par = items$b, 
                              a_par = items$a)
  
  # extract item responses (excluding "subject" column)
  resp <- cog[, c(1:total_items)]
 
  #------------------------------------------------------------------------------#
  # Fit IRT model
  #------------------------------------------------------------------------------#
  # specify the UIRT model
  uirt_fit <- mirt::mirt(resp, 1, itemtype = "2PL", method = "QMCEM", 
                         SE = TRUE, draws = 5000, verbose = FALSE)
  
  uirt_coef <- coef(uirt_fit, printSE = TRUE, IRTpars = T, simplify = T)
  uirt_ip   <- uirt_coef$items

  #------------------------------------------------------------------------------#
  # Summarize and save estimated item parameters
  #------------------------------------------------------------------------------#
  est_ip <- data.frame(test_items,
                       replication = kk, 
                       uirt_ip)

  #------------------------------------------------------------------------------#
  # Scoring
  #------------------------------------------------------------------------------#
  # estimate uirt EAP scores 
  uirt_score <- mirt::fscores( uirt_fit, method = "EAP", full.scores.SE=TRUE) 
  score <- data.frame(  replication = kk,
                        theta,
                        uirt_score )

  #------------------------------------------------------------------------------#
  # Specify the output
  #------------------------------------------------------------------------------#
  output <- list( est_ip = est_ip, score = score )
  return( output )
}

```

After specifying the function, you may distribute a specific replication to a node on the computer.

```{r}
library(parallel) # load the package

cl <- parallel::makeCluster(3) # specify the number of clusters. This is dependent on the number of nodes on the machine you are using.
# serialize the analysis by sending it to multiple cores

sim_output <- parallel::parLapply(cl, replications, myFunction, items = items, test_items = test_items, 
                                  n_examinees = n_examinees, total_items = total_items)    # rename the stored results

# rename each solution of the saved output based on its replication
names(sim_output) <- paste0("r", 1:no_reps)

# save the output to the folder
save(sim_output, file = "sim_output.Rdata")

# stop cluster
stopCluster(cl)

print(proc.time() - ptm) # obtain time
```
We may wish to compare this example with one where we do not parallelize. In this case, all the simulations are randomly specified to one  node simultaneously. 
Using the same test conditions and specification:

```{r}
ptm <- proc.time() # time the analysis

library(mirt)
library(mvtnorm)
library(lsasim)


#------------------------------------------------------------------------------#
# Build container 
#------------------------------------------------------------------------------#
n_examinees   <- 2000 # number of simulated examinees
total_items   <- 30 # total number of items
reps          <- 12 # number of replications

# item_param_reps <- score_reps <-
#   array(NA, 
#         dim = c(reps, total_items), 
#         dimnames = list(1:reps, paste0("i",1:total_items)))

#------------------------------------------------------------------------------#
## Generate item parameters
#------------------------------------------------------------------------------#
items <- list()
set.seed( round(5653) ) # seed for generated item parameters
items <- lsasim::item_gen(n_2pl = total_items , b_bounds = c(-2, 2), 
                          a_bounds = c(0.75, 1.25))

#--- combined sub-test items into a single data frame
test_items <- items

for (aa in 1:reps)
{
  
  set.seed(68575 + 1 + reps)
  #------------------------------------------------------------------------------#
  ## Generate thetas 
  #------------------------------------------------------------------------------#
  theta <- rnorm(n_examinees, mean = 0, sd = 1)
  
  #------------------------------------------------------------------------------#
  # Generate item responses 
  #------------------------------------------------------------------------------#
  resp <- data.frame(matrix(NA, nrow= n_examinees, ncol = total_items))
  colnames(resp) <- paste0("i.", 1:total_items)
  
  # assign items to block
  block_bk1 <- lsasim::block_design(n_blocks = 1, 
                                    item_parameters = items)
  
  #assign block to booklet
  book_bk1 <- lsasim::booklet_design(item_block_assignment = 
                                       block_bk1$block_assignment,
                                     book_design = matrix(1))
  #assign booklet to subjects
  book_samp <- lsasim::booklet_sample(n_subj = n_examinees, 
                                      book_item_design = book_bk1, 
                                      book_prob = NULL)
  
  # generate item responses 
  cog <- lsasim::response_gen(subject = book_samp$subject, 
                              item = book_samp$item, 
                              theta = theta, 
                              b_par = items$b, 
                              a_par = items$a)
  
  # extract item responses (excluding "subject" column)
  resp <- cog[, c(1:total_items)]
  
  #------------------------------------------------------------------------------#
  # Fit IRT model
  #------------------------------------------------------------------------------#
  # specify the UIRT model
  uirt_fit <- mirt::mirt(resp, 1, itemtype = "2PL", method = "QMCEM", 
                         SE = TRUE, draws = 5000, verbose = FALSE)
  
  uirt_coef <- coef(uirt_fit, printSE = TRUE, IRTpars = T, simplify = T)
  uirt_ip   <- uirt_coef$items
  
  #------------------------------------------------------------------------------#
  # Summarize and save estimated item parameters
  #------------------------------------------------------------------------------#
  est_ip <- data.frame(test_items,
                       replication = aa, 
                       uirt_ip)
  save(est_ip, file = paste0("est_ip_", aa, "_reps_", ".Rdata"))
  
  #------------------------------------------------------------------------------#
  # Scoring
  #------------------------------------------------------------------------------#
  # estimate uirt EAP scores 
  uirt_score <- mirt::fscores( uirt_fit, method = "EAP", full.scores.SE=TRUE) 
  score <- data.frame(  replication = aa,
                        theta,
                        uirt_score )
  
  save(score, file = paste0("score_", aa, "_reps_", ".Rdata"))
  
  output <- list( est_ip = est_ip, score = score )
  
  
  
  
}   # close the (aa in 1:reps)  loop

save(output, file = "output.RData")


print(proc.time() - ptm) # obtain time


```
